---
title: "Final_Project_391"
author: "Dorian Ramey"
date: "2026-03-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r cars}
#install.packages("tidytuesdayR")

tuesdata <- tidytuesdayR::tt_load('2025-08-26')

billboard <- tuesdata$billboard
topics <- tuesdata$topics

billboard <- billboard[, c("song", "artist", "date", "cdr_genre", "weeks_at_number_one", "talent_contestant", "keys", "simplified_key", "bpm", "energy", "danceability", "happiness", "length_sec", "lyrical_topic")]

plot(bpm + energy + danceability + happiness + length_sec~date, billboard)
model = lm(weeks_at_number_one~bpm+energy+danceability+happiness+length_sec+simplified_key, billboard)
summary(model)
plot(weeks_at_number_one~date, billboard)
plot(weeks_at_number_one~danceability, billboard)

billboard_talent <- billboard[!is.na(billboard$talent_contestant),] 


unique(billboard$cdr_genre)
table(billboard$cdr_genre)




#split most recent decade, first decade
billboard_recent <- billboard[(billboard$date >= as.Date('2015-01-01')), ]
plot(bpm~date, billboard_recent)

billboard_early <- billboard[(billboard$date <= as.Date('1968-12-31')), ]
plot(bpm~date, billboard_recent)


```

1. What keeps a song at number one? 
Has this changed over time?
--> energy, happiness, date most important factors

2. Can we predict if a song will be a number one hit?

3. Significant world events impact songs at number one

4. Have song factors changed over time? (length, bpm, energy)

```{r features}

#billboard_genre$date <- as.numeric(as.Date(billboard_genre$date) - as.Date("1958-08-04"))

#song length
length <- lm(length_sec~date, data = billboard)
print(summary(length))
plot(length)
plot(length_sec~date, data=billboard, main="Song Length v. Date")

#transformed song length
billboard$date <- billboard$date - 12500
plot(length_sec~abs(date), billboard_genre)
transformed <- lm(length_sec~abs(date), billboard_genre)
print(summary(transformed))

#danceability
dance <- lm(danceability~date, data = billboard)
print(summary(dance))
plot(danceability~date, data=billboard, main="Danceability v. Date")

#energy
energy <- lm(energy~date, data = billboard)
print(summary(energy))
plot(energy~date, data=billboard, main="Energy v. Date")

#bpm
bpm <- lm(bpm~date, data = billboard)
print(summary(bpm))
plot(bpm~date, data=billboard, main="BPM v. Date")


library(stats)
library(dplyr)
library(lubridate)
library(ggplot2)

billboard_genre$cdr_genre_factors <- factor(billboard_genre$cdr_genre)

dance <- lm(as.numeric(cdr_genre_factors)~date, data = billboard_genre)
summary(dance)
plot(dance)

y <- billboard_genre %>% mutate(year = year(date)) %>% group_by(year) %>% summarise(count = n())

y<- as.list(y[, 2])
nrow(y)
length(unique(year(billboard_genre$date)))


billboard_grouped = data.frame(year = numeric(), genre = character(), count = numeric())

for(year in unique(year(billboard_genre$date))){
  for(genre in unique(billboard_genre$cdr_genre)){
    in_year <- billboard_genre[year(billboard_genre$date) == year, ]
    c =  in_year %>% group_by(cdr_genre) %>% summarise(count = n())
    billboard_grouped <- rbind(billboard_grouped, data.frame(year = year, genre = genre, count = c))
  }
}

billboard_grouped$year
library(plotly)
genre_plot <- ggplot(billboard_grouped, aes(x = year, y = count.count, color = count.cdr_genre, text = count.cdr_genre)) + geom_point(size = 3, alpha = 0.7) + labs(x = "Year", y = "Count", title = "Count of Number One Hits of Each Genre per Year")

ggplotly(genre_plot, tooltop = "text")

```
5. What talent shows yield the most number one hits? Are talent shows an indicator of song success? Have talent shows impacted the demographics of singers with a number one hit? (are talent show winners of different race / gender than the average?)
--> hypothesis testing, separate into two groups: null = both groups have same demographics, alt = talent show contestants are of diff demographics

6. How have the most popular genres changed over time? What genres are most or least popular?


# Optional: Stepwise selection starting from a null model
which(is.na(billboard[, c("bpm", "energy", "danceability", "happiness", "length_sec", "weeks_at_number_one", "simplified_key", "cdr_genre")]), arr.ind = TRUE)

billboard <- billboard_tal_inc[c(-691, -835), ]

null_model <- lm(log(weeks_at_number_one) ~ 1, data = billboard)  # Intercept only
full_model <- lm(log(weeks_at_number_one) ~ date + bpm + talent_contestant + energy + danceability + happiness + length_sec + simplified_key, data = billboard, na.action = na.exclude)  # Example full model

# Stepwise selection based on AIC
best_model <- step(null_model, scope = list(lower = null_model, upper = full_model),
                   direction = "forward", trace = FALSE)

cat("\nBest model selected by AIC:\n")
print(summary(best_model))

billboard_genre <- billboard_tal_inc[!is.na(billboard_tal_inc$cdr_genre), ]

for(i in 1:nrow(billboard_genre)){
  if(grepl("Pop", billboard_genre[i,]$cdr_genre)){
    billboard_genre[i, ]$cdr_genre = "Pop"
  } else if(grepl("Rock", billboard_genre[i,]$cdr_genre)){
    billboard_genre[i, ]$cdr_genre = "Rock"
  } else if(grepl("Hip Hop", billboard_genre[i, ]$cdr_genre)){
    billboard_genre[i, ]$cdr_genre = "Hip Hop"
  } else if(grepl("Folk/Country", billboard_genre[i, ]$cdr_genre)){
    billboard_genre[i, ]$cdr_genre = "Folk/Country"
  } else if(grepl("Funk/Soul", billboard_genre[i, ]$cdr_genre)){
    billboard_genre[i, ]$cdr_genre = "Funk/Soul"
  } else if(grepl("Electronic/Dance", billboard_genre[i, ]$cdr_genre)){
    billboard_genre[i, ]$cdr_genre = "Electronic/Dance"
  } 
}


null_model <- lm(log(weeks_at_number_one) ~ 1, data = billboard_genre)  # Intercept only
full_model <- lm(log(weeks_at_number_one) ~ date + bpm + talent_contestant + energy + danceability + happiness + length_sec + simplified_key + cdr_genre, data = billboard_genre, na.action = na.exclude)

# Stepwise selection based on AIC
best_model <- step(null_model, scope = list(lower = null_model, upper = full_model),
                   direction = "both", trace = FALSE)

best_model_forward <- step(null_model, scope = list(lower = null_model, upper = full_model),
                   direction = "forward", trace = FALSE)

cat("\nBest model selected by AIC:\n")
print(summary(best_model_forward))
print(summary(best_model_forward))
print(summary(best_model))
plot(billboard_genre[, c("date", "energy", "happiness", "weeks_at_number_one")])
plot(best_model)
plot(log(weeks_at_number_one) ~ 6.55e-01 + 2.790e-05*date, -2.172e-03*energy, billboard_genre)

billboard_genre$date <- as.numeric(as.Date(billboard_genre$date) - as.Date("1958-08-04"))

plot(weeks_at_number_one ~ (date + energy), billboard_genre)
plot(billboard_genre[, c("date", "weeks_at_number_one", "energy")])

unique(billboard$cdr_genre)


```{r full genre labels}
billboard <- tuesdata$billboard

billboard <- billboard[, c("song", "artist", "date", "cdr_genre", "weeks_at_number_one", "talent_contestant", "keys", "simplified_key", "bpm", "energy", "danceability", "happiness", "length_sec", "lyrical_topic")]

billboard_tal_inc <- billboard

for(i in 1:nrow(billboard_tal_inc)){
  if(is.na(billboard_tal_inc$talent_contestant[i])){
    billboard_tal_inc$talent_contestant[i] = 0
  }else{
    billboard_tal_inc$talent_contestant[i] = 1
  }
}

billboard <- billboard_tal_inc[c(-691, -835), ]

null_model <- lm(log(weeks_at_number_one) ~ 1, data = billboard)  # Intercept only
full_model <- lm(log(weeks_at_number_one) ~ date + bpm + talent_contestant + energy + danceability + happiness + length_sec + simplified_key, data = billboard, na.action = na.exclude)  # Example full model

# Stepwise selection based on AIC
best_model <- step(null_model, scope = list(lower = null_model, upper = full_model),
                   direction = "forward", trace = FALSE)
print(summary(best_model))
```

```{r edited genre labels}
billboard_genre <- billboard_tal_inc[!is.na(billboard_tal_inc$cdr_genre), ]

billboard_genre <- billboard_genre[-which(is.na(billboard_genre)), ]
billboard_genre <- billboard_genre[c(-691, -835), ]

for(i in 1:nrow(billboard_genre)){
  if(grepl("Pop", billboard_genre[i,]$cdr_genre)){
    billboard_genre[i, ]$cdr_genre = "Pop"
  } else if(grepl("Rock", billboard_genre[i,]$cdr_genre)){
    billboard_genre[i, ]$cdr_genre = "Rock"
  } else if(grepl("Hip Hop", billboard_genre[i, ]$cdr_genre)){
    billboard_genre[i, ]$cdr_genre = "Hip Hop"
  } else if(grepl("Folk/Country", billboard_genre[i, ]$cdr_genre)){
    billboard_genre[i, ]$cdr_genre = "Folk/Country"
  } else if(grepl("Funk/Soul", billboard_genre[i, ]$cdr_genre)){
    billboard_genre[i, ]$cdr_genre = "Funk/Soul"
  } else if(grepl("Electronic/Dance", billboard_genre[i, ]$cdr_genre)){
    billboard_genre[i, ]$cdr_genre = "Electronic/Dance"
  } 
}
which(is.na(billboard_genre$bpm))

null_model <- lm(log(weeks_at_number_one) ~ 1, data = billboard_genre, na.action = na.exclude)  # Intercept only

full_model <- lm(log(weeks_at_number_one) ~ date + bpm + talent_contestant + energy + danceability + happiness + length_sec + cdr_genre, data = billboard_genre, na.action = na.exclude)

# Stepwise selection based on AIC
best_model <- step(null_model, scope = list(lower = null_model, upper = full_model),direction = "both", trace = FALSE)

best_model_forward <- step(null_model, scope = list(lower = null_model, upper = full_model), direction = "forward", trace = FALSE)

print(summary(best_model_forward))
print(summary(best_model))

plot(best_model)

```

```{r no genre}
billboard <- billboard[!is.na(billboard$energy),]
null_model <- lm(log(weeks_at_number_one) ~ 1, data = billboard)  # Intercept only
full_model <- lm(log(weeks_at_number_one) ~ date + bpm + talent_contestant + energy + danceability + happiness + length_sec, data = billboard, na.action = na.exclude)

# Stepwise selection based on AIC
best_model <- step(null_model, scope = list(lower = null_model, upper = full_model),direction = "both", trace = FALSE)

best_model_forward <- step(null_model, scope = list(lower = null_model, upper = full_model), direction = "forward", trace = FALSE)

print(summary(best_model_forward))
print(summary(best_model))
plot(billboard_genre[, c("date", "energy", "happiness", "weeks_at_number_one")])
plot(best_model)

```


```{r def genre}
billboard_genre$cdr_genre_factors = factor(billboard_genre$cdr_genre)
null_model <- glm(as.numeric(cdr_genre_factors) ~ 1, data = billboard_genre)  # Intercept only
full_model <- glm(as.numeric(cdr_genre_factors) ~ bpm + energy + danceability + happiness + length_sec + simplified_key, data = billboard_genre, na.action = na.exclude)

# Stepwise selection based on AIC
best_model <- step(null_model, scope = list(lower = null_model, upper = full_model),
                   direction = "both", trace = FALSE)

best_model_forward <- step(null_model, scope = list(lower = null_model, upper = full_model),
                   direction = "forward", trace = FALSE)

print(summary(best_model_forward))
print(summary(best_model))
#plot(billboard_genre[, c("date", "energy", "happiness", "weeks_at_number_one")])
plot(best_model)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
